---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r setoptions, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

Let's load our libraries and data
```{r}
library(ggplot2)
library(lattice)
library(gridExtra)
library(dplyr)
df <- read.csv("activity.csv")
```
  
Now for some initial exploration
```{r}
head(df)
str(df)
dim(df)
```

The date variable is factor, lets convert it to a date. 

```{r}
df$date <- as.Date(df$date, format = "%Y-%m-%d")

```

## What is mean total number of steps taken per day?

```{r}
dfsum <- with(df,tapply(steps, date, sum, na.rm = T))
qplot(dfsum, geom="histogram", binwidth=1000, xlab = "Number of Steps per Day", ylab="Count", main="Histogram for Steps per Day")

cat("Mean Steps per Day", mean(dfsum))
cat("Median Steps per Day", median(dfsum))

```

The mean total number of steps per day is **`r mean(dfsum)`**   
The median total number of steps per day is **`r median(dfsum)`**

## What is the average daily activity pattern?

```{r}
im <- aggregate(steps~interval,data=df,FUN="mean", na.action = na.omit)
ggplot(im, aes(x=interval, y=steps)) + 
    geom_line() +
    labs(x="Interval", y="Average Steps", title="Average Steps per Interval")

cat("Max Interval", im[which.max(im$steps),]$interval)
```
  
The 5 minute interval with most steps is interval **`r im[which.max(im$steps),]$interval`**

## Imputing missing values

```{r}
summary(df)

```

There are 2304 NA's in the steps columns out of 17,568. That is roughly **`r mean(is.na(df$steps))`** or 13%. 


```{r}
head(df[is.na(df$steps),])
```
Lets replace the NAs with the average
```{r}
ndf <- df
ndf[is.na(ndf)] <- mean(df$steps,na.rm=T)
```
With the NA removed lets look at the data again.
```{r}
ndfsum <- with(ndf,tapply(steps, date, sum, na.rm = T))
p1 <- qplot(dfsum, geom="histogram", binwidth=1000, xlab = "Number of Steps per Day", ylab="Count", main="Histogram for Steps per Day")
p2 <- qplot(ndfsum, geom="histogram", binwidth=1000, xlab = "Number of Steps per Day", ylab="Count", main="Histogram for Steps per Day no NAs")
grid.arrange(p1, p2, nrow = 1)

cat("Mean Steps per Day", mean(dfsum), " with no NAs ", mean(ndfsum))
cat("Median Steps per Day", median(dfsum), " with no NAs ", median(ndfsum))

```

The new mean is `r mean(ndfsum)` which is the same as the median `r median(ndfsum)`

## Are there differences in activity patterns between weekdays and weekends?

```{r}
ndf <- tbl_df(ndf)
ndf <- ndf %>% mutate(wkdy = factor(case_when( weekdays(ndf$date) %in% c("Sunday","Saturday") ~ "weekend", TRUE ~ "weekday")))
ndfp <- aggregate(steps~interval + wkdy, data=ndf, FUN="mean")
with(ndfp,xyplot(steps~interval | wkdy, type="l", ylim=c(),scales=list(y=list(relation="free")),layout=c(1,2)))

```